|

### 2023-07-15 paper/03-methods.Rmd
revision edits: strain curves and bootstrap results

Accordingly, the strain rate [@callister2018materials] of a material measures the change in its deformation with respect to time.
Mathematically, the strain rate of a material can be expressed as the derivative of its strain.
That is,
\begin{equation}
\varepsilon'(t) = \frac{d \varepsilon}{dt} = \frac{v(t)}{L_{0}},
\end{equation}
where $v(t)$ is the velocity at which the two points of interest within the material are moving away from or towards each other.
Whereas strain has no units, the strain rate is measured in inverse of time, usually 1/second.

@@

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their team on defense, the opposing team on offense, and their assigned pass blocker.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers matching up with a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the two player groups: pass rushers as $R$ and nearest pass blockers as $B$, as well as for the two team groups: defense $D$ and offense $O$.
We also account for attributes about pass rusher $j$ in play $i$ through the covariate vector $\mathbf{x_{ij}}$, and estimate their respective coefficients $\boldsymbol \beta$ as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + D_{d[i]} + O_{o[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
D_{d} &\sim N(\mu_D, \sigma^2_D), \text{ for } d = 1, \cdots, \text{ \# of teams}, \\
O_{o} &\sim N(\mu_O, \sigma^2_O), \text{ for } o = 1, \cdots, \text{ \# of teams}.
\end{aligned}
\end{equation}

We consider a normal distribution to shrink the random intercepts for each player and team toward their respective group means, which is a useful property this problem's context since we do not observe the same number of plays for each player. For the team effects, this provides us with the average defense and offense team-level effects on a pass rusher's STRAIN. Due to the nested nature of players on teams, the individual pass rusher and blocker random intercepts reflect the respective player's effect relative to their team effects. We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].

With regards to the fixed-effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN. First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects. Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles. These are encoded as indicator variables with defensive ends and tackles as the reference groups for the pass rushers and blockers respectively. We also adjust for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN. Finally, we also adjust for the length of the play (i.e., the elapsed time between the ball snap and a quarterback event). We expect a longer display to display lower STRAIN since this would indicate a quarterback has more time to throw the ball. NEED TO THINK ABOUT THIS MORE....

<!--Additionally, we obtain interval estimates for the player-level random intercepts via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data].-->

@@

In order to provide a measure of uncertainty regarding about our random effects, we use a bootstrapping strategy similar to the approach in @yurko2019nflwar. Specifically, we resample team drives within games. This preserves the fact that team schedules are fixed, but allows for random variation in player usage since this is dependent on team decision making. By resampling plays within the same drive together, our sampling is generating realistic simulated data in comparison to sampling individual plays. For each resampled dataset, we fit the aforementioned multilevel model to generate a distribution of estimates for the considered player and team effects.

>>>>

Accordingly, the strain rate of a material measures the change in its deformation with respect to time.
Mathematically, the strain rate of a material can be expressed as the derivative of its strain.
That is,
\begin{equation}
\varepsilon'(t) = \frac{d \varepsilon}{dt} = \frac{v(t)}{L_{0}},
\end{equation}
where $v(t)$ is the velocity at which the two points of interest within the material are moving away from or towards each other.
Whereas strain has no units, the strain rate is measured in inverse of time, usually 1/second.

@@

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their team on defense, the opposing team on offense, and their assigned pass blocker.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers matching up with a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the two player groups: pass rushers as $R$ and nearest pass blockers as $B$, as well as for the two team groups: defense $D$ and offense $O$.
We also account for attributes about pass rusher $j$ in play $i$ through the covariate vector $\mathbf{x_{ij}}$, and estimate their respective coefficients $\boldsymbol \beta$ as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + D_{d[i]} + O_{o[i]} + \mathbf{x_{ij}} \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
D_{d} &\sim N(\mu_D, \sigma^2_D), \text{ for } d = 1, \cdots, \text{ \# of defensive teams}, \\
O_{o} &\sim N(\mu_O, \sigma^2_O), \text{ for } o = 1, \cdots, \text{ \# of offensive teams}.
\end{aligned}
\end{equation}

We consider a normal distribution to shrink the random intercepts for each player and team toward their respective group means.
This is a useful property since we do not observe the same number of plays for each player. 
For the team effects, this provides us with the average defense and offense team-level effects on a pass rusher's STRAIN.
Due to the nested nature of players on teams, the individual pass rusher and blocker random intercepts reflect the respective player's effect relative to their team effects. 
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].

In order to provide a measure of uncertainty for our random effects, we use a bootstrapping strategy similar to the approach in @yurko2019nflwar.
Specifically, we resample team drives within games, which preserves the fact that team schedules are fixed but allows for random variation in player usage since this is dependent on team decision making.
By resampling plays within the same drive together, our sampling is generating realistic simulated data in comparison to sampling individual plays.
For each bootstrapped dataset, we fit the aforementioned multilevel model to obtain a distribution of estimates for the considered player and team effects.

As for the fixed-effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN.
First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
These are encoded as indicator variables with defensive ends and tackles as the reference levels for the pass rushers and blockers, respectively.
We also adjust for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN.
Finally, we control for play-context covariates with respect to the offensive team.
These include the current down (first, second, third, fourth, or two-point conversion), yards to go for a first down, and current yardline (i.e., distance from the possession team's goal line).
[TODO: add explanation why play context is a proxy for time]

<!-- Finally, we also adjust for the length of the play (i.e., the elapsed time between the ball snap and a quarterback event). We expect a longer display to display lower STRAIN since this would indicate a quarterback has more time to throw the ball. NEED TO THINK ABOUT THIS MORE.... -->

<!-- Additionally, we obtain interval estimates for the player-level random intercepts via parametric bootstrapping of 10,000 samples with the `merTools` package [@knowlesmerTools; @gelman2006data]. -->

@@

### 2023-07-15 paper/04-results.Rmd
revision edits: strain curves and bootstrap results

```{r}
#| label: fig_pos_curves
#| fig.cap: "STRAIN curves for different positions computed by averaging STRAIN values across all players of a particular position and all plays. Edge rushers (OLB and DE) tend to generate more STRAIN than interior rushers (DT and NT). The dotted gray line represents the average STRAIN curve for all players without accounting for position."
#| fig.width: 4.5
#| fig.height: 3
source(here("scripts", "03_fig_pos_curves.R"))
fig_pos_curves
```

@@

Figure \ref{fig:fig_pos_curves} displays the average STRAIN by position for the first 40 frames (4 seconds) after the snap.  For each frame, this curve is calculated by averaging across all players of a particular position (where position is defined by the NFL) and all plays that lasted at least as long as the particular frame of interest meaning that the sample size used to compute these curves diminishes for frames that are chronologically later than frames at the beginning of the play.

We observe a clear difference in STRAIN between edge rushers (OLBs and DEs) and interior linemen (DTs and NTs).
Specifically, edge rushers have higher STRAIN than interior linemen on average, as they are more easily able to approach the quarterback on the edge of the pocket versus, for instance, a nose tackle attacking the line head on.

@@

In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
Figure \ref{fig:fig_rankings_boot} displays the mean varying intercepts for the top ten pass rushers for each position along with the corresponding 95% confidence intervals. 
We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J. Watt (OLB) as the first-ranked pass rusher for their respective position. 
We observe different levels of variability across these top players, with only a subset of players with intervals strictly above zero. This is not necessarily surprising given the limited sample of data.
Additionally, we are ignoring correlated performance between defensive linemen on the same team.
For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), were teammates on the Cleveland Browns during this period of time.
McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an elite pass rusher.
We leave the exploration of adjusting for teammates for future work.

```{r}
#| label: fig_rankings_boot
#| fig.cap: "Top 10 pass rushers for each position (DE, DT, NT, OLB) by mean varying intercept with 95\\% confidence intervals."
#| fig.width: 6
#| fig.height: 8
source(here("scripts", "10_multilevel_mod.R"))
fig_rankings_boot
```

>>>>

```{r}
#| label: fig_pos_curves
#| fig.cap: "STRAIN curves for different positions. Edge rushers (OLB and DE) tend to generate more STRAIN than interior rushers (DT and NT). The dotted gray line represents the average STRAIN curve for all players without accounting for position."
#| fig.width: 4.5
#| fig.height: 3
source(here("scripts", "03_fig_pos_curves.R"))
fig_pos_curves
```

@@

<!-- 
too wordy and confusing imo...
For each frame, this curve is calculated by averaging across all players of a particular position (where position is defined by the NFL) and all plays that lasted at least as long as the particular frame of interest meaning that the sample size used to compute these curves diminishes for frames that are chronologically later than frames at the beginning of the play. -->

Figure \ref{fig:fig_pos_curves} displays the average STRAIN by position for the first 40 frames (4 seconds) after the snap.
For each position, the curve is based on the average of the observed STRAIN at each frame across all plays and players within the position.
We observe a clear difference in STRAIN between edge rushers (OLBs and DEs) and interior linemen (DTs and NTs).
Specifically, edge rushers have higher STRAIN than interior linemen on average, as they are more easily able to approach the quarterback on the edge of the pocket versus, for instance, a nose tackle attacking the line head on.

@@

In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
Figure \ref{fig:fig_rankings_boot} displays the varying intercept distributions for the top ten pass rushers for each position.
We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J. Watt (OLB) as the first-ranked pass rusher for their respective position.
We observe different levels of variability across these top players, with only a subset of players with intercept ranges strictly above zero.
This is not necessarily surprising given the limited sample of data.
Additionally, we are ignoring correlated performance between defensive linemen on the same team.
For instance, the top two defensive ends, Myles Garrett and Takkarist McKinley (see the top panel of Figure \ref{fig:fig_rankings_boot}), were teammates on the Cleveland Browns during this period of time.
McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an elite pass rusher.
We leave the exploration of adjusting for teammates for future work.

```{r}
#| label: fig_rankings_boot
#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (DE, DT, NT, OLB) by median varying intercept"
#| fig.width: 6
#| fig.height: 5
source(here("scripts", "10_multilevel_mod.R"))
fig_rankings_boot
```
